Does freshwater connectivity influence phosphorus retention in lakes?

### 2019-01-20 manuscript/manuscript.Rmd
fill-in citations

liftr:
  maintainer: "Joseph Stachelek"
  email: "stachel2@msu.edu"
  from: "rocker/geospatial:latest"
  pandoc: false
  texlive: true
  sysdeps: 
    - grass
    - p7zip-full
  cran: 
    - LAGOSNE
    - brms
    - dplyr
    - tidyr
    - modelr
    - data.tree
    - ggplot2
    - tidybayes
    - cowplot
  remotes: 
    - "jsta/nhdR"
    - "jsta/spnetwork"
    - "jsta/streamnet"
    - "jsta/rgrass7sf"

@@

Lake phosphorus (P) retention is an important characteristic of lakes that can be used to predict P concentrations and to evaluate lake sensitivity to nutrient loading and eutrophication (Alexander et al. 2008; Milstead et al. 2013). Specifically, P retention is an integrated measure of internal P losses including permanent sedimentation, biological uptake, and other processes that remove P from the water column \citep{chapra2008surface}. P retention has been well studied in lakes because P determines lake trophic status and downstream watershed yields (i.e. export to terminal lakes and coastal estuaries). Although previous studies have shown that lake P retention is related to water residence time \citep{vollenweiderInputoutputModels1975}, large uncertainties exist around this relationship (Brett and Benjamin 2007, Milstead et al. 2013). These uncertainties can be large in lakes with intermediate water residence times, particularly compared to lakes with either extremely short or extremely long water residence times (Figure A1). For example, lakes with very long water residence times (on the order of a decade or longer) have complete or near complete P retention, while lakes with very short water residence times (on the order of days) have almost no P retention (Brett and Benjamin, 2007). The reason for the substantial uncertainty in lake P retention between these two extremes may be that predicting retention solely on the basis of water residence time does not capture many of the other factors and processes that affect P retention (Figure \ref{fig1}).

@@

One well-studied factor that has been shown to influence lake P retention is lake depth through its influence on internal processing of P loads (Søndergaard et al. 2013; Vollenweider 1975). The mechanism for such an influence is that lake depth controls thermal stratification and material resuspension from the benthos. As a result, shallow lakes have a tendency to mix throughout the summer causing redistribution of sedimented phosphorus throughout the mixed zone (Fee et al. 1996). This mixing and redistribution of sedimented P often leads to tighter benthic-pelagic coupling and increased P recycling (Cha et al. 2013). Thus, depth is one example of a lake characteristic that is likely to influence P retention in concert with water residence time (Cheng et al. 2010).

A primary tool for evaluating the influence of specific lake characteristics, like lake depth, on P retention is statistical phosphorus retention modelling. Although the form of such models is variable, many models estimate P retention as a function of water residence time and a parameter k that represents in-lake P decay (Chapra 2008, Vollenweider 1975). Many studies treat k as a constant global parameter (i.e., it has the same value for all lakes), which may be valid only for studies that consider small numbers of lakes in a limited geographic region, with similar characteristics. Few studies have modeled different k values based on lake or watershed characteristics, despite the many differences among lakes that likely influence their ability to process P (Cheng et al. 2010).

For example, there is evidence that the relative proportion of particulate versus dissolved P loads (hereafter, particulate load fraction), influences P retention in stream and wetland ecosystems (Jarvie et al. 2011; Kronvang 1992; Russell et al. 1998; Vanni et al. 2001). However, evidence for particulate load fraction controls on P retention in lakes remains limited because of difficulties in tracking the fate of particulate loads after entering a lake (Brett and Benjamin 2007; Dillon and Molot 1996). Therefore, there is potential to further study particulate load Ffraction using proxies that may be closely related to it, such as the relative amounts of point and nonpoint nutrient sources to a lake subwatershed.

In general, point source inputs to lakes are associated with increased dissolved P loads (Kronvang 1992; Russell et al. 1998), whereas non-point source inputs to lakes, such as those in lake watersheds with high agricultural land use cover, typically have higher particulate P loads (Carpenter et al. 1998; Sharpley et al. 1994). Exceptions to these generalizations have been observed in areas where increased dissolved P loading occurs not as a result of point source nutrient inputs but rather as a result of non-point source runoff due to P saturated soils or legacy P release (Bennett and Carpenter 2001; Powers et al. 2016). Despite the fact that particulate loads are usually related to non-point source inputs, the fraction of non-point source inputs that are in particulate form is highly variable and may depend on intra-annual flow variations as well as watershed erosion characteristics (Jarvie et al. 2011). Furthermore, quantifying non-point source inputs at broad spatial scales is not commonly done due to logistical and sampling constraints (Guy et al. 1994). As a result, studies often infer the relative amounts of particulate and dissolved loading from other available proxy data such as land-use cover (Ellison and Brett 2006, Djodjic and Markensten 2018).

>>>>

liftr:
  maintainer: "Joseph Stachelek"
  email: "stachel2@msu.edu"
  from: "rocker/geospatial:latest"
  pandoc: false
  texlive: true
  sysdeps:
    - grass
    - p7zip-full
  cran:
    - LAGOSNE
    - brms
    - dplyr
    - tidyr
    - modelr
    - data.tree
    - ggplot2
    - tidybayes
    - cowplot
    - nesRdata
    - partykit
    - ggridges
    - pinp
  remotes:
    - "jsta/nhdR"
    - "jsta/spnetwork"
    - "jsta/streamnet"
    - "jsta/rgrass7sf"

@@

Lake phosphorus (P) retention is an important characteristic of lakes that can be used to predict P concentrations and to evaluate lake sensitivity to nutrient loading and eutrophication \citep{alexander2008, milstead2013}. Specifically, P retention is an integrated measure of internal P losses including permanent sedimentation, biological uptake, and other processes that remove P from the water column \citep{chapra2008surface}. P retention has been well studied in lakes because P determines lake trophic status and downstream watershed yields (i.e. export to terminal lakes and coastal estuaries). Although previous studies have shown that lake P retention is related to water residence time \citep{vollenweiderInputoutputModels1975}, large uncertainties exist around this relationship \citep{brett2008, milstead2013}. These uncertainties can be large in lakes with intermediate water residence times, particularly compared to lakes with either extremely short or extremely long water residence times (Figure A1). For example, lakes with very long water residence times (on the order of a decade or longer) have complete or near complete P retention, while lakes with very short water residence times (on the order of days) have almost no P retention \citep{brett2008}. The reason for the substantial uncertainty in lake P retention between these two extremes may be that predicting retention solely on the basis of water residence time does not capture many of the other factors and processes that affect P retention (Figure \ref{fig1}).

@@

One well-studied factor that has been shown to influence lake P retention is lake depth through its influence on internal processing of P loads \citep{sondergaard2013, vollenweiderInputoutputModels1975}. The mechanism for such an influence is that lake depth controls thermal stratification and material resuspension from the benthos. As a result, shallow lakes have a tendency to mix throughout the summer causing redistribution of sedimented phosphorus throughout the mixed zone \citep{fee1996}. This mixing and redistribution of sedimented P often leads to tighter benthic-pelagic coupling and increased P recycling \citep{cha2013}. Thus, depth is one example of a lake characteristic that is likely to influence P retention in concert with water residence time \citep{cheng2010}.

A primary tool for evaluating the influence of specific lake characteristics, like lake depth, on P retention is statistical phosphorus retention modelling. Although the form of such models is variable, many models estimate P retention as a function of water residence time and a parameter k that represents in-lake P decay \citep{chapra2008surface, vollenweiderInputoutputModels1975}. Many studies treat k as a constant global parameter (i.e., it has the same value for all lakes), which may be valid only for studies that consider small numbers of lakes in a limited geographic region, with similar characteristics. Few studies have modeled different k values based on lake or watershed characteristics, despite the many differences among lakes that likely influence their ability to process P \citep{cheng2010}.

For example, there is evidence that the relative proportion of particulate versus dissolved P loads (hereafter, particulate load fraction), influences P retention in stream and wetland ecosystems \citep{jarvie2011, kronvang1992, russell1998, vanni2001}. However, evidence for particulate load fraction controls on P retention in lakes remains limited because of difficulties in tracking the fate of particulate loads after entering a lake \citep{brett2008, dillon1996}. Therefore, there is potential to further study particulate load Ffraction using proxies that may be closely related to it, such as the relative amounts of point and nonpoint nutrient sources to a lake subwatershed.

In general, point source inputs to lakes are associated with increased dissolved P loads \citep{kronvang1992, russell1998}, whereas non-point source inputs to lakes, such as those in lake watersheds with high agricultural land use cover, typically have higher particulate P loads \citep{carpenter1998, sharpley1994}. Exceptions to these generalizations have been observed in areas where increased dissolved P loading occurs not as a result of point source nutrient inputs but rather as a result of non-point source runoff due to P saturated soils or legacy P release \citep{bennet2001, powers2015}. Despite the fact that particulate loads are usually related to non-point source inputs, the fraction of non-point source inputs that are in particulate form is highly variable and may depend on intra-annual flow variations as well as watershed erosion characteristics \citep{jarvie2011}. Furthermore, quantifying non-point source inputs at broad spatial scales is not commonly done due to logistical and sampling constraints \citep{guy1994}. As a result, studies often infer the relative amounts of particulate and dissolved loading from other available proxy data such as land-use cover (Ellison and Brett 2006, Djodjic and Markensten 2018).
